---
title: "Tiny Project: Student Performance and Lifestyle Factors Analysis"
author: "Serafim Ferando Augusto-2505101270238"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Introduction

Student academic performance depends on more than just study effort — family
background, lifestyle habits, and school support all play a role. This Tiny
Project uses a real student survey dataset (395 students, 31 variables) to
explore which factors are associated with whether a student **passed** or
**failed**. The workflow follows the steps covered in Practicals 8, 9 and
10: data importing, preprocessing, statistical exploration, visualization,
and interpretation.

## 2. Data Importing

```{r import}
# Read the CSV file into a data frame.
# stringsAsFactors = FALSE keeps text columns as character for now;
# we convert the relevant ones to factors ourselves in the next section.
student_data <- read.csv("student-data.csv", stringsAsFactors = FALSE)

# Quick look at the data
dim(student_data)      # rows and columns
head(student_data)      # first six records
str(student_data)       # structure and data types of each column
```

**Explanation:** `read.csv()` loads the dataset from the working directory.
`dim()`, `head()` and `str()` give a first impression of the data's size,
sample rows, and column types before any cleaning is done.

## 3. Data Preprocessing and Cleaning

```{r preprocessing}
# Check for missing values in every column
colSums(is.na(student_data))

# Check for duplicate rows
sum(duplicated(student_data))

# Convert key categorical columns to factors for correct summarisation/plots
cat_cols <- c("school", "sex", "address", "famsize", "Pstatus", "Mjob",
              "Fjob", "reason", "guardian", "schoolsup", "famsup", "paid",
              "activities", "nursery", "higher", "internet", "romantic",
              "passed")
student_data[cat_cols] <- lapply(student_data[cat_cols], factor)

# Create one derived feature: average weekly alcohol consumption
# (mean of workday alcohol use 'Dalc' and weekend alcohol use 'Walc')
student_data$alc_avg <- (student_data$Dalc + student_data$Walc) / 2

str(student_data[c(cat_cols, "alc_avg")])
```

**Explanation:** `colSums(is.na(...))` confirms the dataset has no missing
values, and `duplicated()` confirms there are no duplicate rows, so no
imputation or row-dropping is required. The categorical text columns are
converted to `factor` type so R treats them correctly in summaries and
plots. `alc_avg` is a new engineered feature combining weekday and weekend
alcohol consumption into a single lifestyle indicator used later in the
analysis.

## 4. Data Analysis and Statistical Exploration

```{r analysis}
# Overall summary statistics for numeric columns
summary(student_data[, c("age", "studytime", "failures", "absences",
                          "alc_avg")])

# How many students passed vs failed?
table(student_data$passed)

# Pass rate by gender
prop.table(table(student_data$sex, student_data$passed), margin = 1)

# Pass rate by weekly study time category (1=<2hrs, 2=2-5hrs, 3=5-10hrs, 4=>10hrs)
tapply(student_data$passed == "yes", student_data$studytime, mean)

# Average absences for students who passed vs failed
tapply(student_data$absences, student_data$passed, mean)

# Correlation between numeric lifestyle/background variables and absences
num_cols <- c("age", "Medu", "Fedu", "traveltime", "studytime", "failures",
              "famrel", "freetime", "goout", "Dalc", "Walc", "health",
              "absences")
cor(student_data[num_cols])[, "absences"]
```

**Explanation:** `summary()` gives central tendency and spread for the
numeric variables. `table()` and `prop.table()` turn the categorical
`passed` outcome into counts and proportions, split by gender. `tapply()`
computes a statistic (mean) for one variable grouped by another — here, the
pass rate by study-time category, and average absences by pass/fail status.
`cor()` measures the linear relationship between each numeric variable and
absences.

## 5. Graphs and Plots

```{r plot-passed}
result_count <- table(student_data$passed)
pie(result_count,
    main = "Overall Pass vs Fail Distribution",
    labels = paste(names(result_count), result_count))
```

```{r plot-gender}
gender_count <- table(student_data$sex)
barplot(gender_count,
        main = "Gender Distribution of Students",
        xlab = "Gender", ylab = "Number of Students",
        col = c("lightpink", "lightblue"))
```

```{r plot-age}
hist(student_data$age,
     main = "Distribution of Student Age",
     xlab = "Age", col = "lightgreen", breaks = 8)
```

```{r plot-studytime}
pass_rate_by_study <- tapply(student_data$passed == "yes",
                              student_data$studytime, mean) * 100
barplot(pass_rate_by_study,
        main = "Pass Rate (%) by Weekly Study Time Category",
        xlab = "Study Time Category (1=Low ... 4=High)",
        ylab = "Pass Rate (%)", col = "steelblue")
```

```{r plot-absences}
boxplot(absences ~ passed, data = student_data,
        main = "Absences by Pass/Fail Outcome",
        xlab = "Passed", ylab = "Number of Absences",
        col = c("salmon", "lightgreen"))
```

```{r plot-support}
support_rate <- tapply(student_data$passed == "yes",
                        student_data$schoolsup, mean) * 100
barplot(support_rate,
        main = "Pass Rate (%): Extra School Support vs None",
        xlab = "Receives Extra School Support", ylab = "Pass Rate (%)",
        col = "orange")
```

**Explanation:** The `pie()` chart shows the overall pass/fail split. The
`barplot()`s compare categories (gender, study-time groups, school-support
status) against counts or pass rates. `hist()` shows how student ages are
distributed. `boxplot(absences ~ passed, ...)` compares the spread and
median of absences between students who passed and those who failed in one
view.

## 6. Interpretation of Results

- Out of 395 students, **265 (about 67%) passed** and 130 (about 33%)
  failed, so the class is moderately imbalanced toward passing.
- Students who **failed averaged about 6.8 absences**, noticeably higher
  than the **5.2 average absences of students who passed** — attendance
  appears linked to outcome.
- The pass rate rises from roughly **65% for the two lowest study-time
  categories to about 75% and 74%** for the two highest categories,
  suggesting more study time is associated with a better chance of passing,
  though the relationship is not perfectly linear.
- Students who plan to pursue **higher education** had a much higher pass
  rate (about 69%) than those who do not (about 35%) — aspiration/motivation
  appears strongly linked to outcome.
- Interestingly, students **receiving extra school support actually had a
  lower pass rate (about 55%) than those without it (about 69%)**. This is
  a common pattern in observational data: schools likely direct extra
  support toward students who are *already* struggling, so support does not
  cause lower performance — it is a response to it. This shows why
  correlation should not be read as causation.
- Absences correlate most strongly (though still weakly) with **age**
  (r ≈ 0.18) and **weekend alcohol consumption** (r ≈ 0.14) among the
  numeric variables tested, while study time and family relationship
  quality show negligible correlation with absences.

## 7. Findings and Conclusion

This Tiny Project analysed 395 student records to identify factors
associated with academic success. The main findings are:

1. About two-thirds of students in the dataset passed, one-third failed.
2. **Attendance matters** — students who failed had noticeably more
   absences on average than those who passed.
3. **Study time matters, but with diminishing precision** — pass rates
   trend upward with more weekly study time.
4. **Aspiration matters most clearly** — wanting to pursue higher education
   was associated with the largest gap in pass rate of any factor examined.
5. **Extra school support is a marker of prior difficulty, not a cause of
   failure** — it was given to weaker students, so its raw association with
   passing is negative even though its intent is remedial.

Overall, the analysis suggests that consistent attendance, adequate study
time, and student motivation toward further education are the lifestyle
factors most worth encouraging, while formal support programs should be
evaluated by whether they close the gap for the students who receive them,
not by their raw correlation with the pass/fail outcome.
